---
title: "Getting started with the peakable package"
author: "Chao-Jen Wong"
package: peakable
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE,
                      fig.align = 'left')
```

# Getting started
The `peakable` package provides convenient methods for facilitating CUT&RUN and CUT&Tag-seq peaksets quality control and downstream analysis. The functionality breaks down into three parts:     

1. Peak data import/export
2. QC tools including PCA and cosine similarity based on peak hits
3. Finding consensus between replicates
  
## Visit BED and `GRanges`
The BED file is the common format to store peak intervals. The Bioconductor package `rtracklayers` provides tools for importing and exporting stardardized BED+6 and BED+12 files, while `plyranges` offers functions specifically for MACS2 narrow peaks (`read_narrowPeaks()` and `write_narrowpeaks()`). 

The standardized BED+6 columns include _chromosome_, _start_, _end_, _name_, _score_, and _strand_. MACS2 narrow and broad peaks follow this stardard format with additional columns such as _signalValue_, _pValue_, _qValue_, and _peak_ (only for narrow peaks). SEACR peaks, on the other hand, are in the BED+3 format with columns _chromosome_, _start_, and _end_, followed by additional columns _AUC_, _max.signal_, and _max.signal.region_. The `peakable` package provides convenient tools to both costom BED formats. 

In Bioconductor, the imported peaks are formatted as `GRanges` instances, a standard data structure for representing genomic intervals. The `plyranges` pacakge provides `dplyr`-like interface functions for arithematics on `GRanges`. For more information, please visit [here](https://www.bioconductor.org/packages/release/bioc/vignettes/plyranges/inst/doc/an-introduction.html).

# Data import/export

Load libraries:
```{r load-lib}
# load libraries
library(rtracklayer)
library(plyranges)
library(dplyr)
library(ggplot2)

# if peakable not install yet: 
# devtools::install_github('chaochaowong/peakable')
library(peakable)
```
## MACS2 narrow and broad peaks
`read_macs2_narrow()` and `read_macs2_broad()` are wrapper functions to `rtracklayers::import.bed()`, facilitating the import of MACS2 narrow and broad peak files in the BED6+4 format. The imported intervals are formatted as `GRanges` instances.

### narrowPeaks
Import MACS narrowPeaks and exclude non-essential chromosomes such as poorly annotated chromosomes ("chrUn_...") and chrM:
```{r read_macs2_narrow}
# wrapper function of rtracklayer::import.bed()
narrow_file <- 
  system.file('extdata', 
              'chr2_Rep1_H1_CTCF_peaks.narrowPeak', 
              package='peakable')
gr <- read_macs2_narrow(narrow_file,
                        drop_chrM = TRUE,
                        keep_standard_chrom = TRUE,
                        species = 'Homo_sapiens')
gr 
```

Alternatively, `plyranges::read_narrowPeaks()` is also a wrapper function provided by the `plyranges` package to import MACS2 _narrowPeaks_ files.

The `extract_summit_macs2` function extracts the summit from the peak ranges:
```{r extract-macs2-summit}
# extract summit for MACS2 narrowPeaks
summit_macs2 <- extract_summit_macs2(gr)
summit_macs2

# to export a summit GRanges to a bed file, use rtracklayer::export.bed(summit_macs, file=...)
```

To exporting a summit _Granges_ to a bed file, use `rtracklayer::export.bed`. To exporting MACS2 narrowPeaks and preserve the format, use `plyranges::write_narrowPeaks()`.

### broadPeaks
`read_macs2_broad()` imports MACS2 broadPeaks files:

```{r read_macs2_broad}
broad_file <- system.file('extdata', 
                          'chr2_Rep1_H1_H3K27me3_peaks.broadPeak',
                           package='peakable')
gr <- read_macs2_broad(broad_file,
                       drop_chrM = TRUE,
                       keep_standard_chrom = TRUE,
                       species = 'Homo_sapiens')
gr
```

Need to work on `peakable::write_brodPeaks()`.

## SEACR
The `peakable` package provides functions, such as `read_seacr()`, `extract_summit_searcr()`, and `write_seacr()`, for importing and exporting SEACR-specific peak files.

### Import
The metadata columns of SEACR peaks include "AUC", "max.signal", and "max.signal.region" corresponding to peak area under curve, maxinum signal within the peak, and the summit region of the peak.
```{r read_seacr}
seacr_file <- system.file('extdata',
                          'chr2_Rep1_H1_CTCF.stringent.bed',
                          package='peakable')
seacr_gr <- read_seacr(seacr_file,
                 drop_chrM = TRUE,
                 keep_standard_chrom = TRUE,
                 species = 'Homo_sapiens')
seacr_gr
```

The `extract_summit_seacr()` function extracts summit region from the SEACR peaks:
```{r extract-summit-seacr}
summit_seacr <- extract_summit_seacr(seacr_gr)
summit_seacr
```

### Export
The `write_seacr()` function exports the SEACR peaks in SEACR-specific bed format:
```{r write-seacr, eval=FALSE}
write_seacr(seacr_gr, file=file.path('./seacr_peaks.bed'))
```


# Peakset QC

The `peakable` package provides two innovative methods for evaluating the correlation of peak ranges across samples of different antibodies and the similarity between biological repliates.

Both methods involve constructing a peak-hit matix from a collection of peaksets across samples of interest:

__Concepts:__

1. __Constructing a collection of the peaksets of intesrest $C$__: Suppose sample $s$ has a peakset $P_s=\{p_1, p_2, ..., p_l\}$, and there are $m$ peaksets of insterests, $P=\{P_1, P_2, ..., P_s, ..., P_m\}$. The collection of peaksets $C$ is formed by cosolidating and merging nearby ranges in $P$, resulting in a set of $n$ distinct peak ranges.
2. __Building a binary overlap vector for each sample peakset__: For $P_s$, building an overlap vector of size $1 \times n$ with $i$th entry = 1 if any peak range in $P_s$ hits the $i$th range in $C$, or 0 otherwise. 
3. __Building a $n \times m$ peak-hit matrix__: Cancatenating $m$ binary vectors column by column to form an $n-by-m$ peak-hit matrix, where each column corresponds to a sample and each row represents a peak range in $C$.


```{r peak-hit-excalidraw, echo=FALSE, out.width='120%', fig.cap='Peak-hit matrix'}
img <- system.file('images', 'peak-hits.png',
                    package='peakable')
knitr::include_graphics(img)
```

## Peak-based PCA
Once the peak-hit matrix is constructed, we can apply PCA to observe the correlation across peaksets of samples of interest. First, let's contrust a `data.frame` of sample and peak bed files information:


```{r sample-info}
# construct a data.frame of sample information
narrow_pattern = '\\_peaks.narrowPeak$'
sample_info <- data.frame(
  bed_file = list.files(
    system.file('extdata', package = 'peakable'),
    full.names = TRUE, pattern=narrow_pattern)) %>%
  dplyr::mutate(sample_id = 
                  stringr::str_replace(basename(bed_file),
                                       narrow_pattern, ''))   %>%
  dplyr::mutate(antibody = 
                  stringr::str_split(sample_id, '_', 
                                     simplify=TRUE)[, 4]) %>%
  dplyr::relocate(sample_id, antibody, .before='bed_file')

sample_info %>%
  kable(caption='sample information') %>%
  kableExtra::kable_styling('striped')
```

Second, import the peak files as a list of `GRanges`:
```{r import-bed-files}
# import bed files
grl <- lapply(sample_info$bed_file, read_macs2_narrow, 
              drop_chrM = TRUE,
              keep_standard_chrom = TRUE,
              species = 'Homo_sapiens')
names(grl) <- sample_info$sample_id
```

Finally, consolidate the peaks to make the peak-hit matrix (`consolidate_peak_hits()`):

```{r create-hit-matrix-pca}
# construct hit matrix and calculate hit PCA
hits_mat <- peakable::consolidated_peak_hits(grl)
pcs <- peakable:::.getPCA(hits_mat, sample_info, n_pcs=2)
ggplot(pcs, aes(x=PC1, y=PC2, color=antibody)) +
  geom_point() + theme_minimal()
```

## Cosine similarity
The `cos_similarity()` constructs a peak-hits-based matrix of two sets of peaks, with column $u$ and $v$ and apply cos similarity between two binary vectors $\vec{u}$ and $\vec{v}$, i.e., $\frac{\vec{u} \cdot \vec{v}}{\|\vec{u}\|_2 \|\vec{v}\|_2}$.

```{r cos-sim}
# CTCF replicates
ctcf_cos_sim <- cos_similarity(gr_x = grl[["chr2_Rep1_H1_CTCF"]],
                               gr_y = grl[["chr2_Rep2_H1_CTCF"]])

# H3K4me3 replicates
k4me3_cos_sim <- cos_similarity(gr_x = grl[["chr2_Rep1_H1_H3K4me3"]],
                                gr_y = grl[["chr2_Rep2_H1_H3K4me3"]])
```
```{r vis-cos-sim, fig.align='left', fig.height=2, fig.width=6, fig.cap='cos similarity of peak hits between replicates'}
# visualize by ggplot
data.frame(antibody = c('CTCF', 'H3K4me3'),
           cos_sim = c(ctcf_cos_sim, k4me3_cos_sim)) %>%
  ggplot(aes(x=cos_sim, y=antibody)) +
    geom_point() +
    geom_segment(aes(x=0, y=antibody, xend=cos_sim, 
                     yend=antibody), color='grey50') +
    theme_light() +
    labs(title='MACS2 narrow peakets: cos similarity between replicates')
```

# Finding consensus
The `plyranges` package provides many tools, such as `find_overlaps`, `filter_by_overlaps`, to find the overlaps between two peaksets (`GRanges`) while preserving the metadata columns of the peaks. For example:

```{r filter_by_overpals}
x <- grl[['chr2_Rep1_H1_CTCF']]
y <- grl[['chr2_Rep2_H1_CTCF']]
x %>%
  plyranges::filter_by_overlaps(y, minoverlap = 40L)
```

## Consensus for MACS2
`find_consensus_macs2()`
```{r find-consensus-macs2}
consensus <- find_consensus_macs2(x, y, minoverlap = 40L)
consensus
```

Venn diagram of overlaps
```{r find-overlaps-venn}
find_overlaps_venn(x, y, 
                   label_x = 'chr2_Rep1_H1_CTCF',
                   label_y = 'chr2_Rep2_H1_CTCF',
                   minoverlap = 40L)
```

## `consensus_by`

`consensus_by`: take advantage of `sample_info` and robustly get the consensus between replicates
```{r consensus-by}
# consensus_by() group the grl by 'antibody' and returns a data.frame and a list of
# two consensus ranges in GRanges instances
consensus <- 
  peakable:::consensus_by(sample_info, 
                          peaks_grl = grl,
                          consensus_group_by = 'antibody',
                          peak_caller = 'macs2')
consensus$df
head(consensus$grl[['CTCF']])
```
## Consensus for SEACR
`find_consensus_seacr()`
```{r find-consensus-seacr}
# wrapper of plyranges::find_overlaps(); query-based consensus peaks; carry over the metadata
seacr_file_x <- 
  system.file('extdata',
              'chr2_Rep1_H1_CTCF.stringent.bed',
              package='peaklerrr')
seacr_file_y <- 
  system.file('extdata',
              'chr2_Rep2_H1_CTCF.stringent.bed',
              package='peaklerrr')
                            
x <- peaklerrr::read_seacr(seacr_file_x)
y <- peaklerrr::read_seacr(seacr_file_y)
minoverlap <- min(min(width(x)), min(width(y))) / 2
consensus <- find_consensus_seacr(x, y, minoverlap = minoverlap)
consensus

```

# Workflow
`peakable:::peakle_flow()`
